---
title: "Distribution of the Number of Triangles in an Erdös-Rényi RG"
author: Muriel Pérez
date: November 3, 2016
output: md_document
variant: markdown_github
---

output:  pdf_document


#Distribution of the Number of Triangles in an Erdös-Rényi RG

Here I investigate numericaly the limiting distribution of the number
of triangles in a $ER(n,p=\lambda/n)$ graph. I want to confirm that as
the number of nodes $n\to \infty$, the number of triangles
$\Delta_n\to Pois(\lambda^3/6)$ in distribution.

First, I show the R package that I used, igraph.


## igraph

There is a nice R package to work with graphs called
[igraph](http://igraph.org/).

```{r, message = FALSE}
library(igraph)
```

```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
```

igraph is a collection of network analysis tools. Among them, there
are functions to easily sample $ER(n,p=\lambda/n)$ graphs, plot them,
and count the number of triangles, which is what we need. Let's
generate a graph with $n = 100$, $\lambda = 3$, and count the triangles
with the function count_triangles, which returns how many triangles
each vertex is part of.

```{r}
set.seed(2016)
rg <- sample_gnp(n = 100, p = 3/100)
count_triangles(rg) %>%
    sum/3
```

we can also plot rg (See Figure 1).

```{r, fig.cap = "A realization of $ER(n = 100,p=3/100)$"}
plot(rg, vertex.label = NA, vertex.size = 2)
```


## Random sample from $ER(n,p=\lambda/n)$

Now, we want to confirm that as the number of nodes $n\to \infty$, the
number of triangles $\Delta_n\to Poiss(\lambda^3/6)$ in
distribution. For that, I generate a random sample of $N=1000$ random
graphs for $\lambda = 2$, and $n = 10, 20, 100, 200, 500, 1000$, and
then analyze the empirical distribution of $\Delta_{n}$ in each
case. Then I explote what occurs when lambda changes. First, I will
try to use two graphical methods: plotting the probability mass
functions and the qqplot. Then, I will use the $\chi^2$ statistic
using as a reference distribution $Poiss(\lambda^3/6)$ as
evidence of convergence.


```{r echo = FALSE}
generateData <- function(n, lambda, reps){
    reps %>%
        rerun(sample_gnp(n, lambda/n) %>%
              count_triangles %>%
              sum/3) %>%
        unlist
}

calculateFreqs <- function(data, lambda){
    data %>%
        table(dnn = "value") %>%
        as_data_frame %>%
        rename(freq = n) %>%
        mutate(freq = freq/sum(freq),
               value = as.integer(value),
               expected = dpois(value, lambda^3/6))
}

performChisq <- function(data, pois.parameter){
    levs <- data %>%
        factor %>%
        levels %>%
        as.integer
    probs <- dpois(levs, pois.parameter)
    test <- chisq.test(c(table(data),0), p = c(probs, 1-sum(probs)), simulate.p = TRUE)
    return(data_frame(p.value = test$p.value, chisq = test$statistic))
}

lambda <- c(0.1, 2, 3)
n <- c(10, 20, 100, 200, 500, 1000)
reps <- 1000

set.seed(2016)
df <- expand.grid(lambda = lambda, n = n) %>%
    as_data_frame %>%
    mutate(reps = reps) %>%
    mutate(data  = pmap(list(n, lambda, reps), generateData),
           freqs = map2(data, lambda, calculateFreqs),
           test  = map2(data, lambda^3/6, performChisq)) %>%
    unnest(test)

```

We can compare the empirical distribution with the theoretical
probability mass function. For low numbers of $n$, it is obvious that
the distributions do not coincide, while for larger numbers, they get
closer. 

```{r, echo = FALSE, fig.cap = "Observed frequency distribution and expected Poisson distribution"}
## Bar diagrams
df %>%
    filter(n %in% c(10, 100, 1000),
           lambda == 2) %>%
    mutate(n = factor(n)) %>%
    unnest(freqs) %>%
    select(value, freq, expected, n) %>%
    gather(type, p, -c(value,n)) %>%
    ggplot(aes(x = value, y = p)) + geom_bar(aes(fill = type), stat = "identity", position = "dodge") + facet_grid(n~.)
```	

In the same fashion, the qqplots show that for lower numbers of $n$,
the distribution of $\Delta_n$ deviates from the Poisson distribution,
but it gets closer as $n$ grows larger.

```{r, echo = FALSE, fig.cap = "QQ-plots of the sample versus the Poisson distribution quantiles"}
df %>%
    ##filter(n %in% c(10, 20, 100, 1000)) %>%
    filter(lambda==2) %>%
    unnest(data) %>%
    mutate(data = as.integer(data)) %>%
    ggplot(aes(sample = data)) +
    stat_qq(distribution = qpois, dparams = list(lambda = 2^3/6)) +
    geom_abline(slope = 1, intercept = 0) +
    facet_wrap(~n)
```

Finally, we compute the $\chi^2$ for each $n$, and plot it 

```{r, echo = FALSE, fig.cap = "chi^2 statistic as a function of $n$."}
df %>%
    ggplot(aes(x=n, y = chisq, colour = factor(lambda))) + geom_line() + scale_x_log10() + scale_y_log10()
```
